---
title: "Serve Scouting Report"
subtitle: "Cal Poly vs. `r params$opponent`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: false
    self_contained: true
params:
  opponent: "UC Santa Barbara"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(xgboost)

serves    <- readRDS("../data/serves_featured.rds")
quality   <- readRDS("../data/serve_quality.rds")
artifacts <- readRDS("../data/models.rds")
val_metrics <- if (file.exists("../data/validation_metrics.rds")) {
  readRDS("../data/validation_metrics.rds")
} else list(auc = "N/A")

opponent  <- params$opponent
cal_poly  <- "Cal Poly"

# Cal Poly green
cp_green <- "#154734"
```

```{r validate-opponent, include=FALSE}
all_teams <- unique(c(serves$serve_team, serves$opp_team))
if (!opponent %in% all_teams) {
  stop(paste0("Opponent '", opponent, "' not found. Available teams:\n",
              paste(sort(all_teams), collapse = ", ")))
}
```

```{r latest-rates}
# Most recent prior rates per player (from their last recorded contest)
latest_server_rates <- serves %>%
  arrange(contestid) %>%
  group_by(player, serve_team) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(player, serve_team, prior_ace_rate, prior_error_rate,
         prior_fbk_rate, match_ace_rate, match_error_rate)

latest_receiver_rates <- serves %>%
  filter(in_play == 1, !is.na(receiver)) %>%
  arrange(contestid) %>%
  group_by(receiver, opp_team) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(receiver, opp_team, receiver_prior_fbk_rate)

latest_opp_rates <- serves %>%
  filter(in_play == 1) %>%
  arrange(contestid) %>%
  group_by(opp_team) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(opp_team, opp_prior_fbk_rate)
```

---

## Their Weakest Receivers — Serve Targets

Players ranked by historical FBK concession rate (lower = weaker passer = better target).
Minimum 20 receptions. Focus on the top of this list.

```{r receiver-targets}
use_model_rates <- isTRUE(val_metrics$auc_m3 >= 0.55)

receiver_targets <- serves %>%
  filter(in_play == 1, opp_team == opponent, !is.na(receiver)) %>%
  group_by(receiver) %>%
  summarise(
    receptions   = n(),
    raw_fbk_rate = mean(fbk_against),
    .groups      = "drop"
  ) %>%
  filter(receptions >= 20) %>%
  left_join(
    latest_receiver_rates %>% select(receiver, receiver_prior_fbk_rate),
    by = "receiver"
  ) %>%
  mutate(
    fbk_rate = round(
      if (use_model_rates) coalesce(receiver_prior_fbk_rate, raw_fbk_rate) else raw_fbk_rate,
      3
    )
  ) %>%
  arrange(fbk_rate) %>%
  mutate(
    Rank           = row_number(),
    Recommendation = case_when(
      fbk_rate <= quantile(fbk_rate, 0.33) ~ "Primary Target",
      fbk_rate <= quantile(fbk_rate, 0.67) ~ "Secondary Target",
      TRUE                                  ~ "Avoid"
    )
  ) %>%
  select(Rank, Player = receiver, Receptions = receptions, `FBK Rate` = fbk_rate, Recommendation)

kable(receiver_targets, booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 10, latex_options = "hold_position") %>%
  row_spec(which(receiver_targets$Recommendation == "Primary Target"),
           bold = TRUE, color = "white", background = cp_green) %>%
  row_spec(which(receiver_targets$Recommendation == "Avoid"),
           color = "gray")
```

\vspace{6pt}
*FBK Rate = `r if (use_model_rates) "model-adjusted prior rate (receiver_prior_fbk_rate)" else "raw historical rate"`. Lower = weaker passer.*

---

## Their Serving Threats — Who to Watch

Cal Poly's defensive preparation. Opponent servers ranked by Serve Quality Index.
Minimum 20 serves.

```{r serving-threats}
threats <- quality %>%
  filter(serve_team == opponent) %>%
  group_by(player, serve_team) %>%
  summarise(
    Serves   = n(),
    Quality  = round(mean(serve_quality), 3),
    `P(Ace)` = round(mean(p_ace), 3),
    `P(Err)` = round(mean(p_error), 3),
    `P(FBK)` = round(mean(p_fbk), 3),
    .groups  = "drop"
  ) %>%
  filter(Serves >= 20) %>%
  arrange(desc(Quality)) %>%
  mutate(
    Rank             = row_number(),
    `Threat (0-100)` = if (n() > 1) round(100 * (Quality - min(Quality)) / (max(Quality) - min(Quality))) else 100L
  ) %>%
  select(Rank, Player = player, Serves, `Threat (0-100)`, Quality, `P(Ace)`, `P(Err)`, `P(FBK)`)

kable(threats, booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 10, latex_options = "hold_position") %>%
  row_spec(1:min(3, nrow(threats)), bold = TRUE)
```

---

## Cal Poly Serve Matchup

Predicted serve quality for each Cal Poly server against each `r opponent` receiver.
Higher = more favorable matchup for the server. Primary receiver targets highlighted.

```{r matchup-matrix}
# Cal Poly servers with enough volume
cp_servers <- quality %>%
  filter(serve_team == cal_poly) %>%
  group_by(player) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n >= 20) %>%
  pull(player)

# Opponent receivers with enough receptions
opp_receivers <- serves %>%
  filter(in_play == 1, opp_team == opponent, !is.na(receiver)) %>%
  group_by(receiver) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n >= 20) %>%
  pull(receiver)

if (length(cp_servers) == 0 || length(opp_receivers) == 0) {
  cat("Not enough data to generate matchup matrix.")
} else {
  # Build all server × receiver combinations
  matchups <- expand.grid(
    player   = cp_servers,
    receiver = opp_receivers,
    stringsAsFactors = FALSE
  )

  # Attach latest prior rates
  matchups <- matchups %>%
    left_join(latest_server_rates %>% filter(serve_team == cal_poly),
              by = "player") %>%
    left_join(latest_receiver_rates %>% filter(opp_team == opponent),
              by = "receiver") %>%
    mutate(
      opp_prior_fbk_rate = {
        rate <- latest_opp_rates$opp_prior_fbk_rate[latest_opp_rates$opp_team == opponent]
        if (length(rate) == 1) rate else mean(serves$fbk_against[serves$in_play == 1], na.rm = TRUE)
      },
      # Neutral game state
      set_num     = 2L,
      score_diff  = 0L,
      is_home     = 0L,
      is_late_set = 0L,
      # Factor encoding using training levels
      player_id   = as.integer(factor(player,   levels = artifacts$player_levels)),
      receiver_id = as.integer(factor(receiver, levels = artifacts$receiver_levels)),
      opp_team_id = as.integer(factor(opponent, levels = artifacts$opp_levels))
    ) %>%
    mutate(across(c(prior_ace_rate, prior_error_rate, prior_fbk_rate,
                    receiver_prior_fbk_rate, opp_prior_fbk_rate,
                    match_ace_rate, match_error_rate),
                  ~replace(., is.na(.), mean(., na.rm = TRUE))))

  features_fbk  <- artifacts$features_fbk
  features_base <- artifacts$features_base

  X_base <- as.matrix(matchups[, features_base])
  X_fbk  <- as.matrix(matchups[, features_fbk])

  matchups$p_ace     <- predict(artifacts$m1, xgb.DMatrix(X_base))
  matchups$p_error   <- predict(artifacts$m2, xgb.DMatrix(X_base))
  matchups$p_fbk     <- predict(artifacts$m3, xgb.DMatrix(X_fbk))
  matchups$p_in_play <- pmax(0, 1 - matchups$p_ace - matchups$p_error)
  matchups$quality   <- round(matchups$p_ace - matchups$p_error - matchups$p_in_play * matchups$p_fbk, 3)

  # Pivot to wide matrix
  matrix_wide <- matchups %>%
    select(Player = player, receiver, quality) %>%
    tidyr::pivot_wider(names_from = receiver, values_from = quality)

  # Highlight primary targets (top 33% receivers by FBK rate)
  primary_targets <- receiver_targets %>%
    filter(Recommendation == "Primary Target") %>%
    pull(Player)

  tbl <- kable(matrix_wide, booktabs = TRUE, linesep = "",
               caption = "Predicted Serve Quality by Matchup") %>%
    kable_styling(font_size = 9, latex_options = c("scale_down", "landscape"))

  for (col in intersect(primary_targets, colnames(matrix_wide)[-1])) {
    tbl <- tbl %>%
      column_spec(which(colnames(matrix_wide) == col),
                  bold = TRUE, color = cp_green)
  }
  tbl
}
```

---

## Top Recommendations

```{r recommendations}
if (length(cp_servers) > 0 && length(opp_receivers) > 0) {
  top3 <- matchups %>%
    arrange(desc(quality)) %>%
    slice_head(n = 3) %>%
    mutate(
      Rank = row_number(),
      `Serve Quality` = quality,
      `P(Ace)` = round(p_ace, 3),
      `P(FBK)` = round(p_fbk, 3)
    ) %>%
    select(Rank,
           `Cal Poly Server` = player,
           `Target Receiver` = receiver,
           `Serve Quality`,
           `P(Ace)`,
           `P(FBK)`)

  kable(top3, booktabs = TRUE, linesep = "") %>%
    kable_styling(font_size = 10, latex_options = "hold_position") %>%
    row_spec(1, bold = TRUE, color = "white", background = cp_green)
}
```

\vspace{6pt}

```{r recommendation-text, results='asis'}
if (length(cp_servers) > 0 && length(opp_receivers) > 0 && nrow(top3) > 0) {
  best <- top3[1, ]
  cat(paste0(
    "**Primary recommendation:** ", best$`Cal Poly Server`,
    " serving to ", best$`Target Receiver`,
    " (predicted quality: ", best$`Serve Quality`, ", P(FBK) = ", best$`P(FBK)`, ")."
  ))
}
```

\vspace{12pt}
\footnotesize
*Serve Quality = P(Ace) - P(Error) - P(In Play) × P(FBK Against | In Play). All predictions based on historical
player rates and XGBoost model trained on 2025 Big West season PBP data. M3 holdout AUC = `r val_metrics$auc`
(train-only refit on first `r val_metrics$n_train` serves; deployed model trained on full season).
Game state set to neutral (Set 2, tied score, away).*
